---
title: "Silica Analysis with EnvStats"
author: "Lisa Oshita"
date: "June 5, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Ch 5 - Estimating Distribution Quantiles

The following computes the 90th and 95th percentiles, along with upper-bound 90% confidence intervals for those percentiles, of the lognormal distributions estimated from the data we have. 

```{r}
library(EnvStats)

df <- data.frame(silica = c(20, 10, 5, 10, 10, 5),
                 pm10 = c(310, 292, 209, 283, 161, 219))

# 90th percentile with upper confidence limit 
eqlnorm(df$silica, p = 0.90, ci = T, ci.method = "exact", ci.type = "upper",
        conf.level = 0.90)

# 95th percentile with upper confidence limit
eqlnorm(df$silica, p = 0.95, ci = T, ci.method = "exact", ci.type = "upper",
        conf.level = 0.90)
```



# Ch 6 - Prediction Intervals

Prediction intervals for the next k = 1 silica observations, assuming a lognormal distribution.

```{r}
# prediction intervals for lognormal distribution 
df$logsilica <- log(df$silica)

# two-sided prediction limits
predIntNorm(df$logsilica, k = 1, method = "exact", pi.type = "two-sided", conf.level = 0.9)
# lower. upper
exp(1.051401); exp(3.322720)


# upper prediction limit only
predIntNorm(df$logsilica, k = 1, method = "exact", pi.type = "upper", conf.level = 0.9)
# upper
exp(3.018853)

```

* Other prediction intervals:
    + Simultaneous prediction intervals: prediction interval that will contain a specified number of future observations with probability (1 - alpha)100% for each of r future sampling occasions. 



# Ch 6 - Tolerance Intervals

* Definition: an interval on the real line constructed so as to contain B100% of the population (B100% of all future observations). B is called the coverage. 
* B-Content tolerance interval with a 95% confidence level is shown below. This kind of tolerance interval is constructed so that it contains at least 95% of the population. The other form of tolerance interval is B-expectation, which is constructed so that it contains on average B100% of the population. 
    + B-expectation tolerance int is equivalent to a prediction interval for k = 1 future observations. 
    + Note that B-content tolerance intervals will usually be wider than B-expectation TIs. 

```{r}
# upper TI
tolIntLnorm(df$silica, coverage = 0.95, cov.type = "content", ti.type = "upper")

# two-sided TI
tolIntLnorm(df$silica, coverage = 0.95, cov.type = "content", ti.type = "two-sided")
```

## Non-Parametric Tolerance Intervals

* can be used when there is a large amount of non-detects
* based on the ranked data (upper tolerance limits are based on the maximum in the data)

```{r}
# creating a censored variable for the non-detects
df$censor <- FALSE
df$censor[df$silica == 5] <- TRUE

tolIntNpar(df$silica, conf.level = 0.95, ti.type = "upper", lb = 0)
```



# Ch 8 - Censored Data 

## Estimating Distribution Quantiles

The following estimates the quantiles of the lognormal distribution for the data, accounting for the presence of 

```{r}
eqlnormCensored(df$silica, censored = df$censor, censoring.side = "left", 
                p = 0.95, ci = T, ci.method = "normal.approx",
                ci.type = "two-sided", conf.level = 0.95)

# compared to the estimation of quantiles without accounting for censored data
eqlnorm(df$silica, p = 0.95, ci = T, ci.method = "exact", ci.type = "upper",
        conf.level = 0.90)
```

## Prediction Intervals

95% prediction intervals for the next k = 1 silica observations, using methods that account for the presence of censored observations. 

```{r}
enorm.list <- elnormCensored(df$silica, df$censor)
predIntLnorm(enorm.list, 
             k = 1,
             pi.type = "upper",
             conf.level = 0.95, 
             method = "exact")
```



