---
title: "Finding an Evening Threshold"
author: "Lisa Oshita"
date: "November 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(readr)
library(dplyr)
library(lubridate)
```

# Overview

The following investigates the possibility of including an evening threshold for the current text alert algorithm. 

# Load and format the data

```{r load and format data, message=FALSE}
arch.cdf <- read_csv("H:/TECH/Lisa/R/apcd.r/air.alerts/arch-cdf.csv",
                     col_types = list(wsv = "n", wdv = "n", sigt = "n",
                                      bam10 = "n", bam25 = "n", date = "c"))

arch.cdf <- arch.cdf %>%
  mutate(date = parse_date_time(date, "Y!-m!*-d! H!:M!:S!")) %>%
  mutate(day = lubridate::date(date)) %>%
  filter(lubridate::year(date) > 2009)

colnames(arch.cdf)[4] <- "pm"

# format data ---------------------------------------
cdf <- arch.cdf %>%
  mutate(pm = ifelse(pm > 800, NA, pm)) %>%
  mutate(hour = lubridate::hour(date))

# remove days with < 75% completeness
to.remove <- cdf %>%
  group_by(day) %>%
  summarize(prop.complete = 1 - sum(is.na(pm))/24) %>%
  filter(prop.complete < 0.75)

cdf <- arch.cdf %>%
  filter(!day %in% to.remove$day) %>%
  mutate(hour = lubridate::hour(date))

rm(to.remove)
```

# Without evening threshold

```{r}
# initialize functions for testing accuracy
trigger.alert <- function(data, trigger) { # function to determine if alert should be sent out based on trigger
  sapply(data, function(a) sum((a$pm > trigger) & (a$hour > 8), na.rm = TRUE) > 0)
} # added in the restriction that in order for an alert to be triggered, hourly average must be greater than the trigger value at a time later than 8:00am

means <- function(data) {
  sapply(data, function(a) mean(a$pm, na.rm = TRUE)) # function to determine the mean pm for the day
}

acc <- function(pred, true){
  sum(pred == true) / length(true)  
}

fpr <- function(pred, true){
  sum(pred == TRUE & true == FALSE) / length(true)
}

fpr2 <- function(pred, true){
  sum(pred == TRUE & true == FALSE) / sum(pred == TRUE)
}

fnr <- function(pred, true){
  sum(pred == FALSE & true == TRUE) / length(true)
}

fnr2 <- function(pred, true){
  sum(pred == FALSE & true == TRUE) / sum(pred == FALSE)
}

test.trigger <- function(data, trigger, thresh){
  
  true <- means(data) > thresh # days that actually exceeded 24 hour average
  pred <- trigger.alert(data, trigger)
  accuracy <- acc(pred, true)
  false.pos <- fpr(pred, true)
  false.pos2 <- fpr2(pred, true)
  false.neg <- fnr(pred, true)
  false.neg2 <- fnr2(pred, true)
  true.rate <- sum(true)/length(true) # percentage of days that actually exceeded 24 hour average of 50
  
  ret <- c(thresh, true.rate, trigger, accuracy, false.pos, false.pos2, false.neg, false.neg2)
  return(ret)
}
```

```{r}
# split arch.cdf into list by day
cdf.list <- split(cdf, cdf$day)

# run it! 
thresh <- 50
trigs <- c(seq(50, 400, by = 10), 160:180)
results <- test.trigger(cdf.list, 40, thresh) # testing it with just a hourly average trigger of 40

# testing it with a range of triggers
for(i in 1:length(trigs)){
  x <- test.trigger(cdf.list, trigs[i], thresh)
  results <- rbind(results, x)
  rm(x)
} ; rm(i)

results <- as.data.frame(results, row.names = 1:nrow(results))
names(results) <- c("threshold", "true.rate", "trigger", "accuracy", 
                    "false.pos", "false.pos2", "false.neg", "false.neg2")
results <- results[order(results$trigger), ]

# plot it!
plot(results$trigger, results$accuracy, type = "l", ylim = c(0,1),
     main = paste("Target 24-hr average:", thresh, "ug/m3"),
     xlab = "Hourly concentration trigger (ug/m3)",
     ylab = "%",
     yaxt = "n")
axis(2, at = (0:5)/5, labels = (0:5)*20 )
lines(results$trigger, results$false.pos2, col=2)
lines(results$trigger, results$false.pos, col=2, lty=2)
lines(results$trigger, results$false.neg2, col=3)
lines(results$trigger, results$false.neg, col=3, lty=2)

legend("right", 
       legend = c("Accuracy", "False Pos Rate", "Scaled False Pos Rate",
                           "False Neg Rate", "Scaled False Neg Rate"),
       lty = c(1, 2, 1, 2, 1),
       col = c(1, 2, 2, 3, 3),
       bty = "n")

abline(v = results$trigger[which(results$accuracy == max(results$accuracy))[1]],
       lty = 2)
axis(1, 
     at = results$trigger[which(results$accuracy == max(results$accuracy))[1]])
```

# Finding an evening threshold 

* loop over a loop 
* first loop: iterate over time ranges
* second loop: iterate over trigger thresholds (within one iteration of a time range, iterate over different trigger thresholds)

```{r}
trigger.alert.evening.thresh <- function(data, trigger, evening.thresh) { # allows evening hour cutoff to vary
  sapply(data, function(a) sum((a$pm > trigger) & (a$hour > 8) & (a$hour < evening.thresh), na.rm = TRUE) > 0)
}

test.trigger.2 <- function(data, trigger, thresh, evening.thresh){
  true <- means(data) > thresh # days that actually exceeded 24 hour average
  pred <- trigger.alert.evening.thresh(data, trigger, evening.thresh)
  accuracy <- acc(pred, true)
  false.pos <- fpr(pred, true)
  false.pos2 <- fpr2(pred, true)
  false.neg <- fnr(pred, true)
  false.neg2 <- fnr2(pred, true)
  true.rate <- sum(true)/length(true) # percentage of days that actually exceeded 24 hour average of 50
  
  ret <- c(thresh, true.rate, trigger, evening.thresh, accuracy, false.pos, false.pos2, false.neg, false.neg2)
  return(ret)
}


evening.cutoff <- c(23:16) # test by iteratively excluding hours 16 - 23
thresh <- 50
trigs <- c(seq(50, 400, by = 10), 160:180)

results2 <- data.frame(threshold = NA, 
                       true.rate = NA,
                       trigger = NA,
                       evening.cutoff = NA,
                       accuracy = NA,
                       false.pos = NA,
                       false.pos2 = NA,
                       false.neg = NA,
                       false.neg2 = NA)

for (i in evening.cutoff) { 
  for (t in trigs) {
    results2 <- rbind(results2, test.trigger.2(cdf.list, trigger = t, thresh, evening.thresh = i))
  }
}; rm(i, t)

results2 <- results2 %>% arrange(desc(accuracy)) %>% filter(!is.na(threshold))
# trigger value of 179 and an evening cutoff at 17:00/5:00pm resulted in 92% accuracy
```

```{r}
results.for.plot <- results2 %>% filter(evening.cutoff == 17) %>% arrange(trigger)
plot(results.for.plot$trigger, results.for.plot$accuracy, type = "l", ylim = c(0,1),
     main = paste("Target 24-hr average:", thresh, "ug/m3"),
     xlab = "Hourly concentration trigger (ug/m3)",
     ylab = "%",
     yaxt = "n")
axis(2, at = (0:5)/5, labels = (0:5)*20 )
lines(results.for.plot$trigger, results.for.plot$false.pos2, col=2)
lines(results.for.plot$trigger, results.for.plot$false.pos, col=2, lty=2)
lines(results.for.plot$trigger, results.for.plot$false.neg2, col=3)
lines(results.for.plot$trigger, results.for.plot$false.neg, col=3, lty=2)

legend("right", 
       legend = c("Accuracy", "False Pos Rate", "Scaled False Pos Rate",
                           "False Neg Rate", "Scaled False Neg Rate"),
       lty = c(1, 2, 1, 2, 1),
       col = c(1, 2, 2, 3, 3),
       bty = "n")

abline(v = results.for.plot$trigger[which(results.for.plot$accuracy == max(results.for.plot$accuracy))[1]],
       lty = 2)
axis(1, at = results.for.plot$trigger[which(results.for.plot$accuracy == max(results.for.plot$accuracy))[1]])
```


# Comparing algorithms

```{r}
par(mfrow = c(1, 2))
# no evening cutoff
plot(results$trigger, results$accuracy, type = "l", ylim = c(0,1),
     main = paste("No evening cutoff"),
     xlab = "Hourly concentration trigger (ug/m3)",
     ylab = "%",
     yaxt = "n")
axis(2, at = (0:5)/5, labels = (0:5)*20 )
lines(results$trigger, results$false.pos2, col=2)
lines(results$trigger, results$false.pos, col=2, lty=2)
lines(results$trigger, results$false.neg2, col=3)
lines(results$trigger, results$false.neg, col=3, lty=2)

abline(v = results$trigger[which(results$accuracy == max(results$accuracy))[1]],
       lty = 2)
axis(1, 
     at = results$trigger[which(results$accuracy == max(results$accuracy))[1]])

# with evening cutoff of 17
results.for.plot <- results2 %>% filter(evening.cutoff == 17) %>% arrange(trigger)
plot(results.for.plot$trigger, results.for.plot$accuracy, type = "l", ylim = c(0,1),
     main = paste("With an evening cutoff at 17:00"),
     xlab = "Hourly concentration trigger (ug/m3)",
     ylab = "%",
     yaxt = "n")
axis(2, at = (0:5)/5, labels = (0:5)*20 )
lines(results.for.plot$trigger, results.for.plot$false.pos2, col=2)
lines(results.for.plot$trigger, results.for.plot$false.pos, col=2, lty=2)
lines(results.for.plot$trigger, results.for.plot$false.neg2, col=3)
lines(results.for.plot$trigger, results.for.plot$false.neg, col=3, lty=2)

legend("right", 
       legend = c("Accuracy", "False Pos Rate", "Scaled False Pos Rate",
                           "False Neg Rate", "Scaled False Neg Rate"),
       lty = c(1, 2, 1, 2, 1),
       col = c(1, 2, 2, 3, 3),
       bty = "n")

abline(v = results.for.plot$trigger[which(results.for.plot$accuracy == max(results.for.plot$accuracy))[1]],
       lty = 2)
axis(1, at = results.for.plot$trigger[which(results.for.plot$accuracy == max(results.for.plot$accuracy))[1]])
```


```{r}
# without cutoff
knitr::kable(results[which(results$accuracy == max(results$accuracy)), ], format = "html") %>%
  kableExtra::kable_styling(full_width = F) %>%
  kableExtra::column_spec(5, background = "yellow")
```

```{r}
# with cutoff
knitr::kable(results2[which(results2$accuracy == max(results2$accuracy)), ]) %>%
  kableExtra::kable_styling(full_width = F) %>%
  kableExtra::column_spec(5, background = "yellow")
```



