---
title: "Exploring Seasonally Varying PM10 Threshold"
author: "Lisa Oshita"
date: "November 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(readr)
library(dplyr)
library(lubridate)
```

```{r load and format data, message=FALSE}
arch.cdf <- read_csv("H:/TECH/Lisa/R/apcd.r/air.alerts/arch-cdf.csv",
                     col_types = list(wsv = "n", wdv = "n", sigt = "n",
                                      bam10 = "n", bam25 = "n", date = "c"))

arch.cdf <- arch.cdf %>%
  mutate(date = parse_date_time(date, "Y!-m!*-d! H!:M!:S!")) %>%
  mutate(day = lubridate::date(date)) %>%
  filter(lubridate::year(date) > 2009)

colnames(arch.cdf)[4] <- "pm"

# format data ---------------------------------------
cdf <- arch.cdf %>%
  mutate(pm = ifelse(pm > 800, NA, pm)) %>%
  mutate(hour = lubridate::hour(date))

# remove days with < 75% completeness
to.remove <- cdf %>%
  group_by(day) %>%
  summarize(prop.complete = 1 - sum(is.na(pm))/24) %>%
  filter(prop.complete < 0.75)

cdf <- arch.cdf %>%
  filter(!day %in% to.remove$day) %>%
  mutate(hour = lubridate::hour(date))

rm(to.remove)
```

# Plan

* subset the data by month
* perform the same "grid search" as in evening.threshold.Rmd
* determine if there is a big difference in trigger thresholds across months
* if there is - determine the best threshold for each month, then develop new alg that allows threshold to vary by month, and test on all data 
* if no month to month differences - try seasons

```{r}
thresh <- 50
trigs <- c(seq(50, 400, by = 10), 160:180)

monthly <- list()

for (i in 1:12) { 
  cdf.by.month <- cdf %>% filter(lubridate::month(day) == i)
  cdf.split <- split(cdf.by.month, cdf.by.month$day)
  
  monthly.results <- data.frame(threshold = NA, 
                               true.rate = NA,
                               trigger = NA,
                               evening.cutoff = NA,
                               accuracy = NA,
                               false.pos = NA,
                               false.pos2 = NA,
                               false.neg = NA,
                               false.neg2 = NA)
  for (t in trigs) { 
    monthly.results <- rbind(monthly.results, 
                             test.trigger.2(cdf.split, trigger = t, thresh, evening.thresh = 17))
    monthly.results <- monthly.results %>% filter(!is.na(threshold))
  }
  
  monthly[[i]] <- monthly.results
}; rm(i, t)

# finding best threshold for each month
jan <- monthly[[1]]
feb <- monthly[[2]]
mar <- monthly[[3]]
apr <- monthly[[4]]
may <- monthly[[5]]
jun <- monthly[[6]]
jul <- monthly[[7]]
aug <- monthly[[8]]
sept <- monthly[[9]]
oct <- monthly[[10]]
nov <- monthly[[11]]
dec <- monthly[[12]]

results.df <- rbind(jan[which(jan$accuracy == max(jan$accuracy))[1], ],
                    feb[which(feb$accuracy == max(feb$accuracy))[1], ],
                    mar[which(mar$accuracy == max(mar$accuracy))[1], ],
                    apr[which(apr$accuracy == max(apr$accuracy))[1], ],
                    may[which(may$accuracy == max(may$accuracy))[1], ],
                    jun[which(jun$accuracy == max(jun$accuracy))[1], ],
                    jul[which(jul$accuracy == max(jul$accuracy))[1], ],
                    aug[which(aug$accuracy == max(aug$accuracy))[1], ],
                    sept[which(sept$accuracy == max(sept$accuracy))[1], ],
                    oct[which(oct$accuracy == max(oct$accuracy))[1], ],
                    nov[which(nov$accuracy == max(nov$accuracy))[1], ],
                    dec[which(dec$accuracy == max(dec$accuracy))[1], ])

results.df <- cbind(data.frame(month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul",
                                             "Aug", "Sept", "Oct", "Nov", "Dec")),
                        results.df)
```

# Compare results above to just using 175 for all months

```{r}
monthly.175 <- list()
for (i in 1:12) { 
  cdf.by.month <- cdf %>% filter(lubridate::month(day) == i)
  cdf.split <- split(cdf.by.month, cdf.by.month$day)
  
  monthly.results.175 <- data.frame(threshold = NA, 
                               true.rate = NA,
                               trigger = NA,
                               evening.cutoff = NA,
                               accuracy = NA,
                               false.pos = NA,
                               false.pos2 = NA,
                               false.neg = NA,
                               false.neg2 = NA)
  for (t in trigs) { 
    monthly.results.175 <- rbind(monthly.results.175, 
                             test.trigger.2(cdf.split, trigger = 179, thresh, evening.thresh = 17))
    monthly.results.175 <- monthly.results.175 %>% filter(!is.na(threshold))
  }
  
  monthly.175[[i]] <- monthly.results.175
}; rm(i, t)

# finding best threshold for each month
jan.175 <- monthly.175[[1]]
feb.175 <- monthly.175[[2]]
mar.175 <- monthly.175[[3]]
apr.175 <- monthly.175[[4]]
may.175 <- monthly.175[[5]]
jun.175 <- monthly.175[[6]]
jul.175 <- monthly.175[[7]]
aug.175 <- monthly.175[[8]]
sept.175 <- monthly.175[[9]]
oct.175 <- monthly.175[[10]]
nov.175 <- monthly.175[[11]]
dec.175 <- monthly.175[[12]]

results.df.175 <- rbind(jan.175[which(jan.175$accuracy == max(jan.175$accuracy))[1], ],
                    feb.175[which(feb.175$accuracy == max(feb.175$accuracy))[1], ],
                    mar.175[which(mar.175$accuracy == max(mar.175$accuracy))[1], ],
                    apr.175[which(apr.175$accuracy == max(apr.175$accuracy))[1], ],
                    may.175[which(may.175$accuracy == max(may.175$accuracy))[1], ],
                    jun.175[which(jun.175$accuracy == max(jun.175$accuracy))[1], ],
                    jul.175[which(jul.175$accuracy == max(jul.175$accuracy))[1], ],
                    aug.175[which(aug.175$accuracy == max(aug.175$accuracy))[1], ],
                    sept.175[which(sept.175$accuracy == max(sept.175$accuracy))[1], ],
                    oct.175[which(oct.175$accuracy == max(oct.175$accuracy))[1], ],
                    nov.175[which(nov.175$accuracy == max(nov.175$accuracy))[1], ],
                    dec.175[which(dec.175$accuracy == max(dec.175$accuracy))[1], ])

results.df.175 <- cbind(data.frame(month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul",
                                             "Aug", "Sept", "Oct", "Nov", "Dec")),
                        results.df.175)
results.df.175
```

```{r}
compare.df <- data.frame(month = results.df$month, 
                         varying.thresh = results.df$trigger,
                         accuracy.varying.thresh = results.df$accuracy,
                         accuracy.thresh.179 = results.df.175$accuracy) %>%
  mutate(diff.in.accuracy = accuracy.varying.thresh - accuracy.thresh.179)

compare.df
```

# Now trying by seasons

* December to February
* March to June
* July to September
* October to November

```{r}
seasons <- list(c(12, 1, 2), c(3:6), c(7:9), c(10, 11))
seasons.results <- list()

for (i in 1:4) { 
  cdf.by.season <- cdf %>% filter(lubridate::month(day) %in% seasons[[i]])
  cdf.split <- split(cdf.by.season, cdf.by.season$day)
  
  season.res <- data.frame(threshold = NA, 
                           true.rate = NA,
                           trigger = NA,
                           evening.cutoff = NA,
                           accuracy = NA,
                           false.pos = NA,
                           false.pos2 = NA,
                           false.neg = NA,
                           false.neg2 = NA)
  for (t in trigs) { 
    season.res <- rbind(season.res, 
                        test.trigger.2(cdf.split, trigger = t, thresh, evening.thresh = 17))
    season.res <- season.res %>% filter(!is.na(threshold))
  }
  
  seasons.results[[i]] <- season.res
}; rm(i, t, season.res)

winter <- seasons.results[[1]]
spring <- seasons.results[[2]]
summer <- seasons.results[[3]]
fall <- seasons.results[[4]]

results.df.seasons <- rbind(winter[which(winter$accuracy == max(winter$accuracy))[1], ],
                            spring[which(spring$accuracy == max(spring$accuracy))[1], ],
                            summer[which(summer$accuracy == max(summer$accuracy))[1], ],
                            fall[which(fall$accuracy == max(fall$accuracy))[1], ])

# comparing accuracies
compare.df.season <- data.frame(season = c("winter", "spring", "summer", "fall"),
                                varying.thresh = results.df.seasons$trigger,
                                accuracy.varying.thresh = results.df.seasons$accuracy,
                                accuracy.constant.thresh = c(mean(results.df.175$accuracy[c(1, 2, 12)]),
                                                             mean(results.df.175$accuracy[c(3, 4, 5, 6)]),
                                                             mean(results.df.175$accuracy[c(7, 8, 9)]),
                                                             mean(results.df.175$accuracy[c(10, 11)]))) 
compare.df.season <- compare.df.season %>% mutate(diff.in.accuracy = accuracy.varying.thresh - accuracy.constant.thresh)
```

```{r}
compare.df.season
```

```{r}

```












