---
title: "Exploring Seasonally/Monthly-Varying PM10 Thresholds"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(readr)
library(dplyr)
library(lubridate)
```

# Overview

This file investigates the possibility of seasonally/monthly varying PM10 thresholds for the AirAware text alerts. All analysis below is run with an 8:00am - 5:00pm cutoff. 

```{r load and format data, message=FALSE, eval=FALSE}
arch.cdf <- read_csv("H:/TECH/Lisa/R/apcd.r/air.alerts/arch-cdf.csv",
                     col_types = list(wsv = "n", wdv = "n", sigt = "n",
                                      bam10 = "n", bam25 = "n", date = "c"))

arch.cdf <- arch.cdf %>%
  mutate(date = parse_date_time(date, "Y!-m!*-d! H!:M!:S!")) %>%
  mutate(day = lubridate::date(date)) %>%
  filter(lubridate::year(date) > 2009)

colnames(arch.cdf)[4] <- "pm"

# format data ---------------------------------------
cdf <- arch.cdf %>%
  mutate(pm = ifelse(pm > 800, NA, pm)) %>%
  mutate(hour = lubridate::hour(date))

# remove days with < 75% completeness
to.remove <- cdf %>%
  group_by(day) %>%
  summarize(prop.complete = 1 - sum(is.na(pm))/24) %>%
  filter(prop.complete < 0.75)

cdf <- arch.cdf %>%
  filter(!day %in% to.remove$day) %>%
  mutate(hour = lubridate::hour(date))

rm(to.remove)

# NOTE: this file also requires some of the functions from evening.threshold.Rmd
```

```{r load workspace with results}
load("H:/TECH/Lisa/R/apcd.r/air.alerts/varying.threshold.workspace.RData")
```

```{r create functions, eval=FALSE}
# initialize functions for testing accuracy
trigger.alert <- function(data, trigger) { # function to determine if alert should be sent out based on trigger
  sapply(data, function(a) sum((a$pm > trigger) & (a$hour > 8), na.rm = TRUE) > 0)
} # added in the restriction that in order for an alert to be triggered, hourly average must be greater than the trigger value at a time later than 8:00am

means <- function(data) {
  sapply(data, function(a) mean(a$pm, na.rm = TRUE)) # function to determine the mean pm for the day
}

acc <- function(pred, true){
  sum(pred == true) / length(true)  
}

fpr <- function(pred, true){
  sum(pred == TRUE & true == FALSE) / length(true)
}

fpr2 <- function(pred, true){
  sum(pred == TRUE & true == FALSE) / sum(pred == TRUE)
}

fnr <- function(pred, true){
  sum(pred == FALSE & true == TRUE) / length(true)
}

fnr2 <- function(pred, true){
  sum(pred == FALSE & true == TRUE) / sum(pred == FALSE)
}

trigger.alert.evening.thresh <- function(data, trigger, evening.thresh) { # allows evening hour cutoff to vary
  sapply(data, function(a) sum((a$pm > trigger) & (a$hour > 8) & (a$hour < evening.thresh), na.rm = TRUE) > 0)
}

test.trigger.2 <- function(data, trigger, thresh, evening.thresh){
  true <- means(data) > thresh # days that actually exceeded 24 hour average
  pred <- trigger.alert.evening.thresh(data, trigger, evening.thresh)
  accuracy <- acc(pred, true)
  false.pos <- fpr(pred, true)
  false.pos2 <- fpr2(pred, true)
  false.neg <- fnr(pred, true)
  false.neg2 <- fnr2(pred, true)
  true.rate <- sum(true)/length(true) # percentage of days that actually exceeded 24 hour average of 50
  
  ret <- c(thresh, true.rate, trigger, evening.thresh, accuracy, false.pos, false.pos2, false.neg, false.neg2)
  return(ret)
}
```

# Monthly-varying threshold 

```{r, eval=FALSE}
thresh <- 50
trigs <- c(seq(50, 400, by = 10), 160:180)

monthly <- list()

for (i in 1:12) { 
  cdf.by.month <- cdf %>% filter(lubridate::month(day) == i)
  cdf.split <- split(cdf.by.month, cdf.by.month$day)
  
  monthly.results <- data.frame(threshold = NA, 
                               true.rate = NA,
                               trigger = NA,
                               evening.cutoff = NA,
                               accuracy = NA,
                               false.pos = NA,
                               false.pos2 = NA,
                               false.neg = NA,
                               false.neg2 = NA)
  for (t in trigs) { 
    monthly.results <- rbind(monthly.results, 
                             test.trigger.2(cdf.split, trigger = t, thresh, evening.thresh = 17))
    monthly.results <- monthly.results %>% filter(!is.na(threshold))
  }
  
  monthly[[i]] <- monthly.results
}; rm(i, t)

# finding best threshold for each month
jan <- monthly[[1]]
feb <- monthly[[2]]
mar <- monthly[[3]]
apr <- monthly[[4]]
may <- monthly[[5]]
jun <- monthly[[6]]
jul <- monthly[[7]]
aug <- monthly[[8]]
sept <- monthly[[9]]
oct <- monthly[[10]]
nov <- monthly[[11]]
dec <- monthly[[12]]

results.df <- rbind(jan[which(jan$accuracy == max(jan$accuracy))[1], ],
                    feb[which(feb$accuracy == max(feb$accuracy))[1], ],
                    mar[which(mar$accuracy == max(mar$accuracy))[1], ],
                    apr[which(apr$accuracy == max(apr$accuracy))[1], ],
                    may[which(may$accuracy == max(may$accuracy))[1], ],
                    jun[which(jun$accuracy == max(jun$accuracy))[1], ],
                    jul[which(jul$accuracy == max(jul$accuracy))[1], ],
                    aug[which(aug$accuracy == max(aug$accuracy))[1], ],
                    sept[which(sept$accuracy == max(sept$accuracy))[1], ],
                    oct[which(oct$accuracy == max(oct$accuracy))[1], ],
                    nov[which(nov$accuracy == max(nov$accuracy))[1], ],
                    dec[which(dec$accuracy == max(dec$accuracy))[1], ])

results.df <- cbind(data.frame(month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul",
                                             "Aug", "Sept", "Oct", "Nov", "Dec")),
                        results.df)
```

```{r}
knitr::kable(results.df, digits = 3, format = "html") %>%
    kableExtra::kable_styling(full_width = F)
```

```{r, eval=FALSE}
# comparing accuracy of varying threshold to constant threshold of 179
monthly.175 <- list()
for (i in 1:12) { 
  cdf.by.month <- cdf %>% filter(lubridate::month(day) == i)
  cdf.split <- split(cdf.by.month, cdf.by.month$day)
  
  monthly.results.175 <- data.frame(threshold = NA, 
                               true.rate = NA,
                               trigger = NA,
                               evening.cutoff = NA,
                               accuracy = NA,
                               false.pos = NA,
                               false.pos2 = NA,
                               false.neg = NA,
                               false.neg2 = NA)
  for (t in trigs) { 
    monthly.results.175 <- rbind(monthly.results.175, 
                             test.trigger.2(cdf.split, trigger = 179, thresh, evening.thresh = 17))
    monthly.results.175 <- monthly.results.175 %>% filter(!is.na(threshold))
  }
  
  monthly.175[[i]] <- monthly.results.175
}; rm(i, t)

# finding best threshold for each month
jan.175 <- monthly.175[[1]]
feb.175 <- monthly.175[[2]]
mar.175 <- monthly.175[[3]]
apr.175 <- monthly.175[[4]]
may.175 <- monthly.175[[5]]
jun.175 <- monthly.175[[6]]
jul.175 <- monthly.175[[7]]
aug.175 <- monthly.175[[8]]
sept.175 <- monthly.175[[9]]
oct.175 <- monthly.175[[10]]
nov.175 <- monthly.175[[11]]
dec.175 <- monthly.175[[12]]

results.df.175 <- rbind(jan.175[which(jan.175$accuracy == max(jan.175$accuracy))[1], ],
                    feb.175[which(feb.175$accuracy == max(feb.175$accuracy))[1], ],
                    mar.175[which(mar.175$accuracy == max(mar.175$accuracy))[1], ],
                    apr.175[which(apr.175$accuracy == max(apr.175$accuracy))[1], ],
                    may.175[which(may.175$accuracy == max(may.175$accuracy))[1], ],
                    jun.175[which(jun.175$accuracy == max(jun.175$accuracy))[1], ],
                    jul.175[which(jul.175$accuracy == max(jul.175$accuracy))[1], ],
                    aug.175[which(aug.175$accuracy == max(aug.175$accuracy))[1], ],
                    sept.175[which(sept.175$accuracy == max(sept.175$accuracy))[1], ],
                    oct.175[which(oct.175$accuracy == max(oct.175$accuracy))[1], ],
                    nov.175[which(nov.175$accuracy == max(nov.175$accuracy))[1], ],
                    dec.175[which(dec.175$accuracy == max(dec.175$accuracy))[1], ])

results.df.175 <- cbind(data.frame(month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul",
                                             "Aug", "Sept", "Oct", "Nov", "Dec")),
                        results.df.175)
```

```{r, eval=FALSE}
compare.df <- data.frame(month = results.df$month, 
                         varying.thresh = results.df$trigger,
                         accuracy.varying.thresh = results.df$accuracy,
                         accuracy.thresh.179 = results.df.175$accuracy) %>%
  mutate(diff.in.accuracy = accuracy.varying.thresh - accuracy.thresh.179)
colnames(compare.df) <- c("Month", "Monthly.varying.threshold", "Accuracy.varying.thresh", "Accuracy.179.thresh", "Difference.in.accuracy")
```

```{r}
knitr::kable(compare.df, digits = 3,
             caption = "Comparing accuracy of monthly-varying thresholds to accuracy of constant threshold (179)", format = "html") %>%
  kableExtra::kable_styling(full_width = F)
```

# Seasonally-varying threshold

* Winter: December to February
* Spring: March to June
* Summer: July to September
* Fall: October to November

```{r, eval=FALSE}
seasons <- list(c(12, 1, 2), c(3:6), c(7:9), c(10, 11))
seasons.results <- list()

for (i in 1:4) { 
  cdf.by.season <- cdf %>% filter(lubridate::month(day) %in% seasons[[i]])
  cdf.split <- split(cdf.by.season, cdf.by.season$day)
  
  season.res <- data.frame(threshold = NA, 
                           true.rate = NA,
                           trigger = NA,
                           evening.cutoff = NA,
                           accuracy = NA,
                           false.pos = NA,
                           false.pos2 = NA,
                           false.neg = NA,
                           false.neg2 = NA)
  for (t in trigs) { 
    season.res <- rbind(season.res, 
                        test.trigger.2(cdf.split, trigger = t, thresh, evening.thresh = 17))
    season.res <- season.res %>% filter(!is.na(threshold))
  }
  
  seasons.results[[i]] <- season.res
}; rm(i, t, season.res)

winter <- seasons.results[[1]]
spring <- seasons.results[[2]]
summer <- seasons.results[[3]]
fall <- seasons.results[[4]]

results.df.seasons <- rbind(winter[which(winter$accuracy == max(winter$accuracy))[1], ],
                            spring[which(spring$accuracy == max(spring$accuracy))[1], ],
                            summer[which(summer$accuracy == max(summer$accuracy))[1], ],
                            fall[which(fall$accuracy == max(fall$accuracy))[1], ])

# comparing accuracies
compare.df.season <- data.frame(season = c("winter", "spring", "summer", "fall"),
                                varying.thresh = results.df.seasons$trigger,
                                accuracy.varying.thresh = results.df.seasons$accuracy,
                                accuracy.constant.thresh = c(mean(results.df.175$accuracy[c(1, 2, 12)]),
                                                             mean(results.df.175$accuracy[c(3, 4, 5, 6)]),
                                                             mean(results.df.175$accuracy[c(7, 8, 9)]),
                                                             mean(results.df.175$accuracy[c(10, 11)]))) 
compare.df.season <- compare.df.season %>% mutate(diff.in.accuracy = accuracy.varying.thresh - accuracy.constant.thresh)

colnames(compare.df.season) <- c("Season", "Varying.threshold", "Accuracy.varying.thresh", "Accuracy.179.thresh", "Difference.in.accuracy")
```

```{r}
knitr::kable(compare.df.season, digits = 3,
             caption = "Comparing accuracy of seasonally-varying thresholds to accuracy of constant threshold (179)", format = "html") %>%
  kableExtra::kable_styling(full_width = F)
```

# Test thresholds on full data 

### Seasonally-varying thresholds

```{r, eval=FALSE}
evening.cutoff <- 17
average.threshold <- 50

cdf.list <- split(cdf, cdf$day)
day <- cdf.list[[1]]

test.seasonal <- function(data){
  
  # set trigger
  if (month(data$date)[1] %in% c(12, 1, 2)) {trigger <- 270}
  if (month(data$date)[1] %in% c(3, 4, 5, 6)) {trigger <- 170}
  if (month(data$date)[1] %in% c(7, 8, 9)) {trigger <- 178}
  if (month(data$date)[1] %in% c(10, 11)) {trigger <- 230}
  
  # checking if any values are greater than the trigger
  trigger.text <- ifelse(length(which((data$pm >= trigger) & (data$hour > 8) & (data$hour < 17))) > 0, TRUE, FALSE)
  pm10.average <- mean(data$pm, na.rm = TRUE)
  was.correct <- ifelse((trigger.text == TRUE & pm10.average >= 50) | (trigger.text == FALSE & pm10.average < 50), TRUE, FALSE)
  
  
  results.df <- data.frame(date = date(data$date)[1],
                           average.pm10 = pm10.average,
                           text.triggered = trigger.text,
                           was.correct = was.correct)
  
  return(results.df)
  
}

seasonal.results <- purrr::map_dfr(cdf.list, ~test.seasonal(.))
```

```{r}
knitr::kable(head(seasonal.results),
             caption = "Results of the seasonally-varying threshold algorithm for the first 6 days", format = "html") %>%
  kableExtra::kable_styling(full_width = F) 
```

```{r}
# OVERALL ACCURACY
(seasonal.acc <- round(sum(seasonal.results$was.correct)/nrow(seasonal.results), 3))

# FALSE POSITIVE RATE
(seasonal.fpr <- round(sum((seasonal.results$text.triggered == TRUE) & (seasonal.results$was.correct == FALSE)) / nrow(seasonal.results), 3))

# FALSE NEGATIVE RATE
(seasonal.fnn <- round(sum((seasonal.results$text.triggered == FALSE) & (seasonal.results$was.correct == FALSE)) / nrow(seasonal.results), 3))
```

### Monthly-varying thresholds

```{r, eval=FALSE}
test.monthly <- function(data){
  
  # set trigger
  if (month(data$date)[1] == 1) {trigger <- 163}
  if (month(data$date)[1] == 2) {trigger <- 200}
  if (month(data$date)[1] == 3) {trigger <- 170}
  if (month(data$date)[1] == 4) {trigger <- 180}
  if (month(data$date)[1] == 5) {trigger <- 130}
  if (month(data$date)[1] == 6) {trigger <- 171}
  if (month(data$date)[1] == 7) {trigger <- 210}
  if (month(data$date)[1] == 8) {trigger <- 180}
  if (month(data$date)[1] == 9) {trigger <- 160}
  if (month(data$date)[1] == 10) {trigger <- 180}
  if (month(data$date)[1] == 11) {trigger <- 280}
  if (month(data$date)[1] == 12) {trigger <- 220}
  
  # checking if any values are greater than the trigger
  trigger.text <- ifelse(length(which((data$pm >= trigger) & (data$hour > 8) & (data$hour < 17))) > 0, TRUE, FALSE)
  pm10.average <- mean(data$pm, na.rm = TRUE)
  was.correct <- ifelse((trigger.text == TRUE & pm10.average >= 50) | (trigger.text == FALSE & pm10.average < 50), TRUE, FALSE)
  
  
  results.df <- data.frame(date = date(data$date)[1],
                           average.pm10 = pm10.average,
                           text.triggered = trigger.text,
                           was.correct = was.correct)
  
  return(results.df)
  
}

monthly.results <- purrr::map_dfr(cdf.list, ~test.monthly(.))
```

```{r}
knitr::kable(head(monthly.results),
             caption = "Results of the monthly-varying threshold algorithm for the first 6 days", format = "html") %>%
  kableExtra::kable_styling(full_width = F) 
```

```{r}
# OVERALL ACCURACY
(monthly.acc <- round(sum(monthly.results$was.correct)/nrow(monthly.results), 3))

# FALSE POSITIVE RATE
(monthly.fpr <- round(sum((monthly.results$text.triggered == TRUE) & (monthly.results$was.correct == FALSE)) / nrow(monthly.results), 3))

# FALSE NEGATIVE RATE
(monthly.fnn <- round(sum((monthly.results$text.triggered == FALSE) & (monthly.results$was.correct == FALSE)) / nrow(monthly.results), 3))
```

### Constant threshold (179)

```{r, eval=FALSE}
test.constant.thresh <- function(data){
  
  # checking if any values are greater than the trigger
  trigger.text <- ifelse(length(which((data$pm >= 179) & (data$hour > 8) & (data$hour < 17))) > 0, TRUE, FALSE)
  pm10.average <- mean(data$pm, na.rm = TRUE)
  was.correct <- ifelse((trigger.text == TRUE & pm10.average >= 50) | (trigger.text == FALSE & pm10.average < 50), TRUE, FALSE)
  
  
  results.df <- data.frame(date = date(data$date)[1],
                           average.pm10 = pm10.average,
                           text.triggered = trigger.text,
                           was.correct = was.correct)
  
  return(results.df)
  
}

constant.thresh.results <- purrr::map_dfr(cdf.list, ~test.constant.thresh(.)) 
```

```{r}
# OVERALL ACCURACY
(constant.acc <- round(sum(constant.thresh.results$was.correct)/nrow(constant.thresh.results), 3))

# FALSE POSITIVE RATE
(constant.fpr <- round(sum((constant.thresh.results$text.triggered == TRUE) & 
                             (constant.thresh.results$was.correct == FALSE)) / nrow(constant.thresh.results), 3))

# FALSE NEGATIVE RATE
(constant.fnn <- round(sum((constant.thresh.results$text.triggered == FALSE) & 
                             (constant.thresh.results$was.correct == FALSE)) / nrow(constant.thresh.results), 3))
```

# Comparing thresholds

```{r, eval=FALSE}
comparisons <- data.frame(Constant.threshold = c(constant.acc, constant.fpr, constant.fnn),
                          Monthly.varying = c(monthly.acc, monthly.fpr, monthly.fnn),
                          Seasonally.varying = c(seasonal.acc, seasonal.fpr, seasonal.fnn))

rownames(comparisons) <- c("Accuracy", "False positive rate", "False negative rate")
```

```{r}
knitr::kable(comparisons,
             caption = "Comparing accuracy, false positive rate, false negative rate across the different thresholds", format = "html") %>%
  kableExtra::kable_styling(full_width = F) 
```




